---
title: "Plots clustering on ISCs"
author: "Jordi Camps"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r message=FALSE}
library(SingleCellExperiment)
library(scater)
library(scTools)
library(edgeR)
library(ggthemes)
library(ggExtra)
library(SC3)
library(ggplot2)
library(dplyr)
library(scran)
library(viridis)
library(RColorBrewer)
set.seed(1234567)
```

## Load clustering file
```{r}
sce <- readRDS("data/reads_qc_scran_ISC_SC3.rds")
sce
```

## Define colors
```{r}
library(RColorBrewer)
col_k3 <- brewer.pal(3, "Set2")
#col_k5 <- brewer.pal(5, "Set1")
col_k5 <- c("#66c2a5", "#fc8d62", "#76BAFF", "#785ECC", "#AD54B2")
#alternative colors: "#55B270", "#FF9B55",
col_genotype <- c("lightslategrey", "firebrick3")
```

## Plot clusters on PCA
```{r fig.height=4, fig.width=5}
reducedDim(sce) <- NULL
plotPCA(sce, ncomponents = 4, colour_by = "sc3_3_clusters")
plotPCA(sce, ncomponents = 4, colour_by = "sc3_4_clusters")
plotPCA(sce, ncomponents = 4, colour_by = "sc3_5_clusters")
```

## QC on  clusters
### Load functions
```{r}
make_col_ann_for_heatmaps <- function(object, show_pdata) {
    if (any(!show_pdata %in% colnames(colData(object)))) {
        show_pdata_excl <- show_pdata[!show_pdata %in% colnames(colData(object))]
        show_pdata <- show_pdata[show_pdata %in% colnames(colData(object))]
        message(paste0("Provided columns '", paste(show_pdata_excl, collapse = "', '"), "' do not exist in the phenoData table!"))
        if (length(show_pdata) == 0) {
            return(NULL)
        }
    }
    ann <- NULL
    if (is.null(metadata(object)$sc3$svm_train_inds)) {
        ann <- colData(object)[, colnames(colData(object)) %in% show_pdata]
    } else {
        ann <- colData(object)[metadata(object)$sc3$svm_train_inds, colnames(colData(object)) %in% 
            show_pdata]
    }
    # remove columns with 1 value only
    if (length(show_pdata) > 1) {
        keep <- unlist(lapply(ann, function(x) {
            length(unique(x))
        })) > 1
        if (!all(keep)) {
            message(paste0("Columns '", paste(names(keep)[!keep], collapse = "', '"), "' were excluded from annotation since they contained only a single value."))
        }
        ann <- ann[, names(keep)[keep]]
        if (ncol(ann) == 0) {
            ann <- NULL
        } else {
            ann <- as.data.frame(lapply(ann, function(x) {
                if (nlevels(as.factor(x)) > 9) 
                  x else as.factor(x)
            }))
            # convert outlier scores back to numeric
            for (i in grep("_log2_outlier_score", colnames(ann))) {
                if (class(ann[, i]) == "factor") {
                  ann[, i] <- as.numeric(levels(ann[, i]))[ann[, i]]
                }
            }
        }
    } else {
        if (length(unique(ann)) > 1) {
            ann <- as.data.frame(ann)
            colnames(ann) <- show_pdata
            if (!grepl("_log2_outlier_score", show_pdata)) {
                ann <- as.data.frame(lapply(ann, function(x) {
                  if (nlevels(as.factor(x)) > 9) 
                    return(x) else return(as.factor(x))
                }))
            }
        } else {
            message(paste0("Column '", show_pdata, "' was excluded from annotation since they contained only a single value."))
            ann <- NULL
        }
    }
    return(ann)
}
```

### Consensus plot
```{r fig.height=5, fig.width=7}
#png("plots/QC/consensus_k7.png", width = 6, height = 5, res = 600)
k = 5
object = sce
show_pdata = "sc3_5_clusters"

    hc <- metadata(object)$sc3$consensus[[as.character(k)]]$hc
    #hc <- dendextend::rotate(hc, c(256:145))
    consensus <- metadata(object)$sc3$consensus[[as.character(k)]]$consensus
    
    add_ann_col <- FALSE
    ann <- NULL
    if (!is.null(show_pdata)) {
        ann <- make_col_ann_for_heatmaps(object, show_pdata)
        if (!is.null(ann)) {
            add_ann_col <- TRUE
            # make same names for the annotation table
            rownames(ann) <- colnames(consensus)
        }
    mat_colors <- list(sc3_5_clusters = col_k5)
    names(mat_colors$sc3_5_clusters) <- unique(sce$sc3_5_clusters)[order(unique(sce$sc3_5_clusters))]
    }
    pheatmap::pheatmap(mat = consensus, color = colorRampPalette(rev(brewer.pal(n = 5, name =
"RdYlBu")))(100), cluster_rows = hc, cluster_cols = hc, show_rownames = FALSE, show_colnames = FALSE, treeheight_row = 0, treeheight_col = 0, annotation_names_col = FALSE, legend_breaks = c(0, 1), annotation_col = ann[add_ann_col], annotation_colors = mat_colors, fontsize = 14, annotation_legend = FALSE, filename = "plots/QC/ISC/consensus_k5.tiff", width = 4, height = 4)
```

Cluster stability
```{r}
SC3::sc3_plot_cluster_stability(object = sce, k = 5)
```

Silhouette plot
```{r fig.height=5, fig.width=6}
pdf("plots/QC/ISC/silhouette_k5.pdf", width = 3.5, height = 3.5)
plot(metadata(sce)$sc3$consensus$`5`$silhouette, col = col_k5)
dev.off()
```

Export PCA dims
```{r fig.height=4, fig.width=5}
reducedDim(sce) <- NULL
sce = plotPCA(sce, ncomponents = 4, colour_by = "sc3_5_clusters", return_SCE = TRUE)
sce$PC1 <- reducedDim(sce)[,1]
sce$PC2 <- reducedDim(sce)[,2]
sce$PC3 <- reducedDim(sce)[,3]
sce$PC4 <- reducedDim(sce)[,4]
```

PCA on genotype
```{r fig.height=3, fig.width=2.5}
#genotype by PC1
plot_dims(sce_object = sce, x = "PC1", y = "PC2", color = "genotype", point_size = 3, alpha = 1, theme = 14) +
  #labs(x = "Principal component 1", y = "Principal component 2") +
  scale_color_manual("changed legend", values = c(Healthy = "#4e79a7", Dystrophic = "#f28e2b"), 
                       labels = c(Healthy = "Healthy", Dystrophic = "Dystrophic")) +
  #scale_shape_manual("changed legend", values = c(Healthy = 16, Dystrophic = 17), labels = c(Healthy = "Healthy", Dystrophic = "Dystrophic")) +
  guides(shape = FALSE) +
  theme(legend.position = "top", legend.title = element_blank())
ggsave("plots/ISC/ISC_PCA_genotype.pdf", dpi = 600, width = 2.5, height = 3)
ggsave("plots/ISC/ISC_PCA_genotype.tiff", dpi = 600, width = 2.5, height = 3)
```

### ISC1, ISC2 and ISC3
```{r fig.height=2.5, fig.width=3.5}
plot_dims(sce_object = sce, x = "PC1", y = "PC2", color = "Cluster", point_size = 3, alpha = 1, theme = 14) +
  theme(legend.position = "right", legend.title = element_blank()) +
  scale_color_brewer(type = "qual", palette = "Set2") +
  guides(shape = FALSE) +
ggsave("plots/ISC/ISC1_2_3.pdf", dpi = 600, width = 3.5, height = 2.5)
ggsave("plots/ISC/ISC1_2_3.tiff", dpi = 600, width = 3.5, height = 2.5)
```

### PCA on k=5
Set names k5
```{r}
sce$sc3_5_clusters <- plyr::revalue(sce$sc3_5_clusters, c("1" = "ISC 3c", "2" = "ISC 3a", "3" = "ISC 3b", "4" = "ISC 1", "5" = "ISC 2"))
sce$sc3_5_clusters <- factor(sce$sc3_5_clusters, levels = c("ISC 1", "ISC 2", "ISC 3a", "ISC 3b", "ISC 3c"))
```


PC1-2
```{r fig.height=2, fig.width=3}
#healthy clusters by PC2
plot_dims(sce_object = sce, x = "PC1", y = "PC2", color = "sc3_5_clusters", point_size = 3, alpha = 1, theme = 12) +
  theme(legend.position = "right", legend.title = element_blank(), axis.line = element_line(size = 1, colour = "black")) +
  #scale_color_brewer(type = "qual", palette = "Set1") +
  scale_color_manual(values = col_k5) +
  guides(shape = FALSE)
ggsave("plots/ISC/ISC_PCA_k5_PC1_PC2.pdf", dpi = 600)
ggsave("plots/ISC/ISC_PCA_k5_PC1_PC2.tiff", dpi = 600)
```

PC1-3
```{r fig.height=2.5, fig.width=3.5}
#Dystrophic clusters by PC3
plot_dims(sce_object = sce, x = "PC1", y = "PC3", color = "sc3_5_clusters", point_size = 3, alpha = 1, theme = 14) +
  #scale_color_brewer(type = "qual", palette = "Set1") +
  scale_color_manual(values = col_k5) +
  theme(legend.position = "right", legend.title = element_blank()) +
  guides(shape = FALSE)
ggsave("plots/ISC/ISC_PCA_k5_PC1_PC3.pdf", dpi = 600, width = 3.5, height = 2.5)
ggsave("plots/ISC/ISC_PCA_k5_PC1_PC3.tiff", dpi = 600, width = 3.5, height = 2.5)
```

PC2-3
```{r fig.height=2.5, fig.width=3.5}
#Healthy and dystrophic by PC2 and PC3
plot_dims(sce_object = sce, x = "PC2", y = "PC3", color = "sc3_5_clusters", point_size = 3, alpha = 1, theme = 14) +
  #scale_color_brewer(type = "qual", palette = "Set1") +
  scale_color_manual(values = col_k5) +
  theme(legend.position = "right", legend.title = element_blank()) +
  guides(shape = FALSE)
ggsave("plots/ISC/ISC_PCA_k5_PC2_PC3.pdf", dpi = 600, width = 3.5, height = 2.5)
ggsave("plots/ISC/ISC_PCA_k5_PC2_PC3.tiff", dpi = 600, width = 3.5, height = 2.5)
```

PC1-4
```{r fig.height=2.5, fig.width=3.5}
#Mixed cluster by PC4
plot_dims(sce_object = sce, x = "PC1", y = "PC4", color = "sc3_5_clusters", point_size = 3, alpha = 1, theme = 14) +
  #scale_color_brewer(type = "qual", palette = "Set1") +
  scale_color_manual(values = col_k5) +
  theme(legend.position = "right", legend.title = element_blank()) +
  guides(shape = FALSE)
ggsave("plots/ISC/ISC_PCA_k5_PC1_PC4.pdf", dpi = 600, width = 3.5, height = 2.5)
ggsave("plots/ISC/ISC_PCA_k5_PC1_PC4.tiff", dpi = 600, width = 3.5, height = 2.5)
```

### 3D scatterplot
```{r eval=FALSE, include=FALSE}
require(rgl)
require(RColorBrewer)
plot3d(sce$PC1, sce$PC2, sce$PC3, xlab = "Principal component 1", ylab = "Principal component 2", zlab = "Principal component 3",
       col = brewer.pal(5, "Set1")[unclass(sce$sc3_5_clusters)], size = 12)
```

genotype per cluster
```{r fig.height=2.75, fig.width=2.25}
#sce$sc3_5_clusters <- factor(sce$sc3_5_clusters, levels = c(1, 3, 2, 5, 4))
library(scales)
colData(sce) %>%
  as_tibble() %>%
  ggplot(aes(x = sc3_5_clusters, fill = genotype)) +
  geom_bar(aes(y = (..count..)/sum(..count..)), position = "fill", width = 0.8, alpha = 1) +
  #geom_hline(yintercept = 0.5, size = .5, linetype = 2, alpha = 0.6) +
  #coord_flip() +
  scale_fill_manual(values = col_genotype) +
  scale_y_continuous("Percentage per population", breaks = c(0.25, 0.5, 0.75, 1), expand = c(0, 0), labels = c("25%", "50%", "75%", "100%")) +
  scale_x_discrete("Clustering on ISC") +
  #guides(fill = guide_legend(keyheight = 0.8, keywidth = 0.8)) +
  #labs(fill = "Clustering\non Lin-") +
  theme_bw(base_size = 12) +
  theme(panel.border = element_blank(), axis.line = element_line(colour = "black", size = 1), axis.ticks.x = element_blank(),
        panel.grid = element_blank(), axis.ticks.y = element_line(colour = "black", size = 1), legend.title = element_blank(),
        axis.text = element_text(colour = "black"), legend.position = "top", axis.title.x = element_blank(), 
        axis.text.x = element_text(angle = 90, vjust = 0.3, hjust = 1), axis.title.y = element_text(hjust = .5))
ggsave("plots/ISC/fraction_genotype_k5.pdf", dpi = 600)
ggsave("plots/ISC/fraction_genotype_k5.tiff", dpi = 600)
```

cluster per genotype
```{r fig.height=2.5, fig.width=3}
#sce$sc3_5_clusters <- factor(sce$sc3_5_clusters, levels = c(1, 3, 2, 5, 4))
library(scales)
colData(sce) %>%
  as_tibble() %>%
  ggplot(aes(x = genotype, fill = sc3_5_clusters)) +
  geom_bar(aes(y = (..count..)/sum(..count..)), position = "fill", width = 0.7) +
  #geom_hline(yintercept = 0.5, size = .5, linetype = 2, alpha = 0.6) +
  #coord_flip() +
  scale_fill_manual(values = col_k5) +
  scale_y_continuous("Percentage per population", breaks = c(0.25, 0.5, 0.75, 1), expand = c(0, 0), labels = c("25%", "50%", "75%", "100%")) +
  scale_x_discrete("Cluster") +
  guides(fill = guide_legend(keyheight = 0.7, keywidth = 0.7)) +
  theme_bw(base_size = 12) +
  theme(panel.border = element_blank(), axis.line = element_line(colour = "black", size = 0.5), axis.ticks.x = element_blank(),
        panel.grid = element_blank(), axis.title.x = element_blank(), axis.ticks.y = element_line(colour = "black", size = 0.5),
        axis.text = element_text(colour = "black"), legend.title = element_blank(), legend.position = "right", 
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("plots/ISC/fraction_clusters_per_genotype_k5.pdf", dpi = 600, width = 3.5, height = 2.5)
ggsave("plots/ISC/fraction_clusters_per_genotype_k5.tiff", dpi = 600, width = 3.5, height = 2.5)
```

ISC 1, 2 and 3 in k=5
```{r fig.height=2.25, fig.width=3.25}
library(scales)
colData(sce) %>%
  as_tibble() %>%
  ggplot(aes(fill = sc3_5_clusters, x = Cluster)) +
  geom_bar(aes(y = ..count..), width = 0.8) +
  #geom_hline(yintercept = 0.5, size = 1.5, linetype = 3) +
  #coord_flip() +
  scale_fill_manual(values = col_k5) +
  scale_y_continuous("Number of cells per cluster", expand = c(0, 0), breaks = c(25, 50, 75)) +
  scale_x_discrete("Clustering on Lin-") +
  #guides(fill = guide_legend(keyheight = 0.8, keywidth = 0.8)) +
  labs(fill = "Clustering\non ISC") +
  theme_bw(base_size = 12) +
  theme(panel.border = element_blank(), axis.line = element_line(colour = "black", size = 1), axis.ticks.x = element_blank(),
        panel.grid = element_blank(), axis.ticks.y = element_line(colour = "black", size = 1),
        axis.text = element_text(colour = "black"), legend.position = "right", axis.title.y = element_text(hjust = 1))
ggsave("plots/ISC/fraction_ISC1-3_in_k5.pdf", dpi = 600)
ggsave("plots/ISC/fraction_ISC1-3_in_k5.tiff", dpi = 600)
```


Genes for dystrophic clusters
```{r fig.height=2.5, fig.width=2.5}
plot_expression(sce, var = c("Pparg", "Fabp4", "Adipoq"), group = "Cluster", facet = "vertical", theme = 14) +
  stat_summary(fun.y = median, geom = "point", size = 2) +
  scale_fill_manual(values = col_k3) +
  scale_color_manual(values = col_k3) +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 0.5), strip.text.x = element_text(size = 12),
        axis.ticks.y = element_line(colour = "black", size = 0.5), axis.text.y = element_text(colour = "black", size = 10), 
        axis.text.x = element_text(colour = "black", size = 11, angle = 45, hjust = 1))
ggsave("plots/ISC/adipo_genes.pdf", dpi = 600, height = 2.5, width = 2.5)
```




### Plot genes per cluster with violin plot
#### Markers for ISC2
```{r fig.height=3, fig.width=3}
plot_expression(sce, var = c("Gdf10", "Cpe", "F3"), group = "sc3_5_clusters", facet = "vertical", theme = 12, scale = "width") +
  stat_summary(fun.y = median, geom = "point", size = 2) +
  scale_fill_manual(values = col_k5) +
  scale_color_manual(values = col_k5) +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 0.5), strip.text.x = element_text(size = 12),
        axis.ticks.y = element_line(colour = "black", size = 0.5), axis.text.y = element_text(colour = "black", size = 8), 
        axis.text.x = element_text(colour = "black", size = 10, angle = 45, hjust = 1))
ggsave("plots/ISC/markers_ISC2_k5.pdf", dpi = 600, width = 3, height = 3)
ggsave("plots/ISC/markers_ISC2_k5.tiff", dpi = 600, width = 3, height = 3)
```

```{r fig.height=3, fig.width=3}
plot_expression(sce, var = c("Gdf10", "Cpe", "F3"), group = "Cluster", facet = "vertical", theme = 12) +
  stat_summary(fun.y = median, geom = "point", size = 2) +
  scale_fill_manual(values = col_k3) +
  scale_color_manual(values = col_k3) +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 0.5), strip.text.x = element_text(size = 14),
        axis.ticks.y = element_line(colour = "black", size = 0.5), axis.text = element_text(colour = "black", size = 8), 
        axis.text.x = element_text(colour = "black", size = 12, angle = 45, hjust = 1))
ggsave("plots/ISC/markers_ISC2_k3.pdf", dpi = 600, width = 3, height = 3)
ggsave("plots/ISC/markers_ISC2_k3.tiff", dpi = 600, width = 3, height = 3)
```


#### ISC1 markers
```{r fig.height=3, fig.width=3}
plot_expression(sce, var = c("Cd55", "Anxa3", "F3"), group = "sc3_5_clusters", facet = "vertical", theme = 12) +
  stat_summary(fun.y = median, geom = "point", size = 2) +
  scale_fill_manual(values = col_k5) +
  scale_color_manual(values = col_k5) +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 0.5), strip.text.x = element_text(size = 14),
        axis.ticks.y = element_line(colour = "black", size = 0.5), axis.text.y = element_text(colour = "black", size = 8), 
        axis.text.x = element_text(colour = "black", size = 10, angle = 45, hjust = 1))
ggsave("plots/ISC/markers_ISC1_k5.pdf", dpi = 600, width = 3, height = 3)
ggsave("plots/ISC/markers_ISC1_k5.tiff", dpi = 600, width = 3, height = 3)
```

```{r fig.height=3, fig.width=3}
plot_expression(sce, var = c("Cd55", "Anxa3"), group = "Cluster", facet = "vertical", theme = 12) +
  stat_summary(fun.y = median, geom = "point", size = 2) +
  scale_fill_manual(values = col_k3) +
  scale_color_manual(values = col_k3) +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 0.5), strip.text.x = element_text(size = 14),
        axis.ticks.y = element_line(colour = "black", size = 0.5), axis.text.y = element_text(colour = "black", size = 8), 
        axis.text.x = element_text(colour = "black", size = 10, angle = 45, hjust = 1))
ggsave("plots/ISC/markers_ISC1.pdf", dpi = 600, width = 3, height = 3)
ggsave("plots/ISC/markers_ISC1.tiff", dpi = 600, width = 3, height = 3)
```

#### Markers ISC3
```{r fig.height=3, fig.width=3}
plot_expression(sce, var = c("Thbs4", "Fbln7", "Sdc1"), group = "sc3_5_clusters", facet = "vertical", theme = 12) +
  stat_summary(fun.y = median, geom = "point", size = 2) +
  scale_fill_manual(values = col_k5) +
  scale_color_manual(values = col_k5) +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 0.5), strip.text.x = element_text(size = 14),
        axis.ticks.y = element_line(colour = "black", size = 0.5), axis.text.y = element_text(colour = "black", size = 8),
        axis.text.x = element_text(colour = "black", size = 10, angle = 45, hjust = 1))
ggsave("plots/ISC/markers_ISC3_k5.pdf", dpi = 600, width = 3, height = 3)
ggsave("plots/ISC/markers_ISC3_k5.tiff", dpi = 600, width = 3, height = 3)
```

```{r fig.height=3, fig.width=3}
plot_expression(sce, var = c("Thbs4", "Fbln7", "Sdc1"), group = "Cluster", facet = "vertical", theme = 12) +
  stat_summary(fun.y = median, geom = "point", size = 2) +
  scale_fill_manual(values = col_k3) +
  scale_color_manual(values = col_k3) +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 0.5), strip.text.x = element_text(size = 14),
        axis.ticks.y = element_line(colour = "black", size = 0.5), axis.text.y = element_text(colour = "black", size = 8),
        axis.text.x = element_text(colour = "black", size = 10, angle = 45, hjust = 1))
ggsave("plots/ISC/markers_ISC3.pdf", dpi = 600, width = 3, height = 3)
ggsave("plots/ISC/markers_ISC3.tiff", dpi = 600, width = 3, height = 3)
```

### Markers ISC3b and ISC3c
```{r fig.height=2.5, fig.width=3}
plot_expression(sce, var = c("G0s2", "Ptx3", "Cthrc1"), group = "sc3_5_clusters", facet = "vertical", theme = 12) +
  stat_summary(fun.y = median, geom = "point", size = 2) +
  scale_fill_manual(values = col_k5) +
  scale_color_manual(values = col_k5) +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 1),x`
        axis.ticks.y = element_line(colour = "black", size = 1), axis.text.y = element_text(colour = "black"),
        axis.text.x = element_text(colour = "black", angle = 45, hjust = 1))
ggsave("plots/ISC/markers_ISC3b-c.pdf", dpi = 600)
ggsave("plots/ISC/markers_ISC3b-c.tiff", dpi = 600)
```


#### ISC markers per genotype
```{r fig.height=4, fig.width=5}
sce_object = sce
gene = c("Cd55", "F3", "Sdc1")
group = "sc3_5_clusters"

rowData <- NULL
rowData  <- t(logcounts(sce_object)[gene, ])
rowData <- data.frame(rowData)
  colData <- data.frame(cluster = sce_object[[group]], genotype = sce_object[["genotype"]])
  temp <- cbind(rowData, colData)
  temp <- tidyr::gather(temp, gene, logcounts, -cluster, -genotype)
  #temp$gene <- factor(temp$gene, levels = var)
  
ggplot(temp, aes(cluster, logcounts)) +
  geom_boxplot(aes(fill = cluster)) +
  facet_grid(gene ~ genotype)
ggsave("plots/ISC/markers_per_cluster_and_genotype.pdf", dpi = 600, width = 5, height = 4)



ggplot(temp, aes(cluster, logcounts)) +
  geom_jitter(aes(col = cluster), width = 0.5) +
  facet_grid(gene~genotype)
```


#### Print markers per gene
```{r fig.height=2.5, fig.width=3.5, include=FALSE}
markers <- c("Gdf10", "Thbd", "F3", "Cd55", "Ly6a", "Anxa3", "Thbs4", "Fbln7", "Sdc1", "Cthrc1", "H19", "Ptx3",  "Ier5", "G0s2", "Tsc22d3", "Pdgfra")

for (m in markers) {
  plot_expression(sce, var = m, group = "Cluster", facet = "horizontal", theme = 14, type = "boxplot") +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = .75), strip.text.x = element_text(size = 16),
        axis.ticks.y = element_line(colour = "black", size = .75), axis.text = element_text(colour = "black"))
ggsave(paste("plots/ISC/", m, ".pdf", sep = ""), dpi = 600, width = 2.25, height = 2.5)
ggsave(paste("plots/ISC/", m, ".tiff", sep = ""), dpi = 600, width = 2.25, height = 2.5)
}
```

```{r fig.height=3, fig.width=4, include=FALSE}
markers <- c("Gdf10", "Thbd", "F3", "Cd55", "Ly6a", "Anxa3", "Thbs4", "Fbln7", "Sdc1", "Cthrc1", "H19", "Ptx3",  "Ier5", "G0s2", "Tsc22d3", "Pdgfra")

for (m in markers) {
  plot_expression(sce, var = m, group = "sc3_5_clusters", facet = "horizontal", theme = 14, type = "boxplot") +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  scale_y_continuous("Log(expression)", expand = c(0, 0)) +
  theme(panel.border = element_blank(), strip.background = element_blank(), legend.position = "none", 
        axis.line = element_line(colour = "black", size = 0.5), strip.text.x = element_text(size = 20),
        axis.ticks.y = element_line(colour = "black", size = 0.5), axis.text = element_text(colour = "black"))
ggsave(paste("plots/ISC/", m, "k5.pdf", sep = ""), dpi = 600, width = 3, height = 3.5)
ggsave(paste("plots/ISC/", m, "k5.tiff", sep = ""), dpi = 600, width = 3, height = 3.5)
}
```

Statistics on markers
```{r}
markers <- c("Cd55", "F3", "Sdc1")
```

Make dataframe with genes and expression
```{r}
group = "Cluster"

  rowData <- NULL
  temp <- NULL
  for (gene in markers) {
    if (gene %in% row.names(sce)) {
      rowData[[gene]]  <- logcounts(sce)[gene, ]
    }
    else {
      print(paste0(gene, " not expressed or written wrong!", sep = ""))
    }

  rowData <- data.frame(rowData)
  colData <- data.frame(cluster = sce[[group]])
  temp <- cbind(rowData, colData)
  temp <- tidyr::gather(temp, gene, logcounts, -cluster)
temp$gene <- factor(temp$gene, levels = markers)
}
```

```{r}
var_a <- c("ISC 1", "ISC 2", "ISC 3")
var_b <- c("ISC 2", "ISC 3", "ISC 1")


for (m in markers) {
  temp2 <- temp %>% 
  filter(gene == m)
  print(paste("Wilcoxon test ", m, sep = ""))
  for (i in 1:3) {
    temp3 <- temp2 %>%
      filter(cluster == var_a[i] | cluster == var_b[i])
    test <- wilcox.test(temp3$logcounts~temp3$cluster)
    print(paste(var_a[i], "vs", var_b[i], "p =", test$p.value, sep = " "))
  }
}
```


### Correlation between Gdf10 and surface markers
```{r}
rowData <- t(logcounts(sce)[c("Gdf10", "F3"), ])
rowData <- as.data.frame(rowData)
rowData$cluster <- sce$Cluster
rowData$genotype <- sce$genotype

ggplot(as_tibble(rowData), aes(x = Gdf10, y = F3, col = cluster, shape = genotype)) +
  geom_point(size = 3) +
  scale_color_brewer(type = "qual", palette = "Set2") +
  scale_x_continuous(breaks = c(0, 10)) +
  scale_y_continuous(breaks = c(0, 10)) +
  #theme_bw(base_size = 18) +
  geom_rangeframe(col = "black") +
  theme_tufte(base_size = 18) +
  theme(axis.text = element_text(colour = "black"))
```


### tSNE
Try out different perplexities
```{r fig.height=4, fig.width=5.5}
for (p in 4:10) {
  print(plotTSNE(sce, colour_by = "sc3_5_clusters", perplexity = p, exprs_values = "logcounts", rand_seed = 123456, return_SCE = FALSE) + ggtitle(paste("Perplexity = ", p, sep = "")))
}
```

Export tSNE dims
```{r fig.height=4, fig.width=5.5}
reducedDims(sce)$TSNE <- NULL
plotTSNE(sce, colour_by = "sc3_5_clusters", perplexity = 5, exprs_values = "logcounts", rand_seed = 123456, return_SCE = FALSE)
sce <- plotTSNE(sce, colour_by = "sc3_5_clusters", return_SCE = TRUE, perplexity = 5, exprs_values = "logcounts", rand_seed = 123456)
sce$Dim1 <- reducedDims(sce)$TSNE[,1]
sce$Dim2 <- reducedDims(sce)$TSNE[,2]
```

Plot tSNE k=5
```{r fig.height=2.5, fig.width=3.5}
plot_dims(sce, x = "Dim1", y = "Dim2", color = "sc3_5_clusters", theme = 14, point_size = 3) +
  scale_color_manual(values = col_k5) +
  labs(x = "Dim 1", y = "Dim 2") +
  guides(shape = FALSE) +
  theme(legend.position = "right", legend.title = element_blank())
ggsave("plots/ISC/tSNE_k5.pdf", dpi = 600, width = 3.5, height = 2.5)
```

