---
title: "Scenic"
author: "Jordi Camps"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dir.create("int")
dir.create("output")
options(stringsAsFactors = FALSE)
library(BiocGenerics)
library(SingleCellExperiment)
```

# Step 1: inferring potential TF targets based on co-expression
## Load and filter data
### Load expression matrix
```{r}
sce <- readRDS("data/reads_qc_scran_sc3.rds")
exprMat <- counts(sce)
dim(exprMat)
```

### Save cell information
```{r}
cellInfo <- colData(sce)[colnames(exprMat), "sc3_4_clusters", drop = F]
colVars <- list(sc3_4_clusters=setNames(c("forestgreen", "darkorange", "magenta4", "hotpink"), 
                                        c("FAP/MAB", "Schwann cells", "Satellite cells", "Smooth muscle cells")))
save(colVars, file="data/colVars.Rdata")
plot.new(); legend(0,1, fill=colVars$sc3_4_clusters, legend=names(colVars$sc3_4_clusters))
```

### Select organism
```{r}
org <- "mm9"
if(org=="hg19")
{
  library(RcisTarget.hg19.motifDatabases.20k)
  
  # Get genes in databases:
  data(hg19_500bpUpstream_motifRanking) # or 10kbp, they should have the same genes
  genesInDatabase <- hg19_500bpUpstream_motifRanking@rankings$rn
  
  # Get TFS in databases:
  data(hg19_direct_motifAnnotation)
  allTFs <- hg19_direct_motifAnnotation$allTFs
}

if(org=="mm9")
{
  library(RcisTarget.mm9.motifDatabases.20k)
  
  # Get genes in databases:
  data(mm9_500bpUpstream_motifRanking) # or 10kbp, they should have the same genes
  genesInDatabase <- mm9_500bpUpstream_motifRanking@rankings$rn
  
  # Get TFS in databases:
  data(mm9_direct_motifAnnotation)
  allTFs <- mm9_direct_motifAnnotation$allTFs
}
```

### Gene filter/selection
Dataset already filtered aforehand
```{r}
nCellsPerGene <- apply(exprMat, 1, function(x) sum(x>0))
nCountsPerGene <- apply(exprMat, 1, sum)
summary(nCellsPerGene)
summary(nCountsPerGene)
max(exprMat)
sum(exprMat>0) / sum(exprMat==0)
```

Remove genes with less than 70 counts (equivalent of 20 counts expressed in 1% of the total cells)
```{r}
minReads <- 20*.01*ncol(exprMat)
genesLeft_minReads <- names(nCountsPerGene)[which(nCountsPerGene > minReads)]
length(genesLeft_minReads)
```


Remove genes that are epxressed in less than 1% of the cells
```{r}
minSamples <- ncol(exprMat)*.01
nCellsPerGene2 <- nCellsPerGene[genesLeft_minReads]
genesLeft_minCells <- names(nCellsPerGene2)[which(nCellsPerGene2 > minSamples)]
length(genesLeft_minCells)
```

```{r}
genesLeft_minCells_inDatabases <- genesLeft_minCells[which(genesLeft_minCells %in% genesInDatabase)]
length(genesLeft_minCells_inDatabases)
```

Filter expression matrix
```{r}
exprMatrix_filtered <- exprMat[genesLeft_minCells_inDatabases, ]
save(exprMatrix_filtered, file="int/1.1_exprMatrix_filtered.RData")
```

Check whether any relevant gene/potential gene of interest is missing:
```{r}
interestingGenes <- c("Pax7", "Sox10", "Ly6a", "Gdf10", "Alpl")
interestingGenes[which(!interestingGenes %in% rownames(exprMatrix_filtered))]
```

```{r}
rm(exprMat)
```

### Potential regulators: list of transcription factors
```{r}
inputTFs <- allTFs[allTFs%in% rownames(exprMatrix_filtered)]
save(inputTFs, file="int/1.2_inputTFs.RData")

c(allTFs=length(allTFs), inputTFs=length(inputTFs))
```

## GENIE3
```{r}
# setwd("")
load("int/1.1_exprMatrix_filtered.RData")
# Optional: add log (if it is not logged already)
exprMatrix_filtered <- log2(exprMatrix_filtered+1) 
load("int/1.2_inputTFs.RData")
library(GENIE3)
```

### Run GENIE3
```{r}
set.seed(123)
weightMatrix <- GENIE3(exprMatrix_filtered, regulators=inputTFs, nCores=4)
save(weightMatrix, file="int/1.3_GENIE3_weightMatrix.RData")
```

### Correlation
```{r}
load("int/1.1_exprMatrix_filtered.RData")
corrMat <- cor(t(exprMatrix_filtered), method="spearman")
save(corrMat, file="int/1.4_corrMat.RData")
```

## Create co-expression modules
### Load output from GENIE3
```{r}
library(GENIE3)
# Convert the weight matrix into links:
load("int/1.3_GENIE3_weightMatrix.RData")
linkList <- getLinkList(weightMatrix, threshold=0.001) # (slighly faster)
# linkList <- getLinkList(weightMatrix)
colnames(linkList) <- c("TF", "Target", "weight")
# order by weight
linkList <- linkList[order(linkList[,"weight"], decreasing=TRUE),]
save(linkList, file="int/1.5_GENIE3_linkList.RData")
```

```{r}
load("int/1.5_GENIE3_linkList.RData")
dim(linkList)
head(linkList)
```

### Creating TF-modules (potential TF-targets)
#### Building gene-sets
```{r}
quantile(linkList$weight, probs=c(0.75, 0.90))
plot(linkList$weight[1:1000000], type="l", ylim=c(0, max(linkList$weight)), main="Weight of the links",
     ylab="Weight", xlab="Links sorted decreasingly")
abline(h=0.001, col="blue") # Threshold
sum(linkList$weight>0.001)/nrow(linkList)
```

```{r}
linkList_001 <- linkList[which(linkList[,"weight"]>0.001),]
# Number of links over the threshold: 
nrow(linkList_001) 
```

Create gene-sets and save
```{r}
tfModules <- list()

linkList_001$TF <- as.character(linkList_001$TF)
linkList_001$Target <- as.character(linkList_001$Target)
  
#### Create TF-modules:
# 1: Weight > 0.001 (filtered in previous step) 
tfModules[["w001"]] <- split(linkList_001$Target, factor(linkList_001$TF))

# 2: Weight > 0.005
llminW <- linkList_001[which(linkList_001[,"weight"]>0.005),]
tfModules[["w005"]] <- split(llminW$Target, factor(llminW$TF))

# 3: Top 50 targets for each TF
# ("w001" should be ordered decreasingly by weight)
tfModules[["top50"]] <- lapply(tfModules[["w001"]], function(x) x[1:(min(length(x), 50))])

# 4-6: Top regulators per target 
# (linkList_001 should be ordered by weight!)
linkList_001_byTarget <- split(linkList_001, factor(linkList_001$Target))
save(linkList_001_byTarget, file="int/1.5_linkList_001_byTarget.RData")

nTopTfs <- c(5, 10, 50)
nTopTfs <- setNames(nTopTfs, paste("top", nTopTfs, "perTarget", sep=""))

library(reshape2); library(data.table)
topTFsperTarget <- lapply(linkList_001_byTarget, function(llt) {
   nTFs <- nTopTfs[which(nTopTfs <= nrow(llt))]
   melt(lapply(nTFs, function(x) llt[1:x,"TF"]))
})
topTFsperTarget <- topTFsperTarget[which(!sapply(sapply(topTFsperTarget, nrow), is.null))]
topTFsperTarget.asDf <-  data.frame(rbindlist(topTFsperTarget, idcol=TRUE))
head(topTFsperTarget.asDf)
colnames(topTFsperTarget.asDf) <- c("Target", "TF", "method")

# Merge the all the gene-sets:
tfModules.melted <- melt(tfModules)
colnames(tfModules.melted) <- c("Target", "TF", "method")
tfModules <- rbind(tfModules.melted, topTFsperTarget.asDf)

save(tfModules, file="int/1.6_tfModules.RData")
```

```{r}
load("int/1.6_tfModules.RData")
# Basic counts:
rbind(nGeneSets=nrow(tfModules), 
      nTFs=length(unique(tfModules$TF)), 
      nTargets=length(unique(tfModules$Target)))
```

#### Split into positive- and negative-correlated targets
Split TF targets according to correlation
```{r}
load("int/1.4_corrMat.RData")
# Keep only correlation between TFs and potential targets
tfs <- unique(tfModules$TF)
corrMat <- corrMat[tfs,]

# Split TF modules according to correlation
tfModules_byTF <- split(tfModules, factor(tfModules$TF))
tfModules_withCorr_byTF <- lapply(tfModules_byTF, function(tfGeneSets)
{
    tf <- unique(tfGeneSets$TF)
    targets <- tfGeneSets$Target
    cbind(tfGeneSets, corr=c(as.numeric(corrMat[tf,targets] > 0.03) - as.numeric(corrMat[tf,targets] < -0.03)))
})
tfModules_withCorr <- data.frame(rbindlist(tfModules_withCorr_byTF))
save(tfModules_withCorr, file="int/1.7_tfModules_withCorr.RData")
```

```{r}
load("int/1.7_tfModules_withCorr.RData")
head(tfModules_withCorr)
dim(tfModules_withCorr)
```

# Step 2: identifying regulons (direct TF targets) based on DNA motif enrichment
## Load gene sets
```{r}
load("int/1.7_tfModules_withCorr.RData")

# Remove genes missing from RcisTarget databases
#  (In case the input matrix wasn't already filtered)
tfModules_withCorr <- tfModules_withCorr[which(as.character(tfModules_withCorr$TF) %in% allTFs),]
geneInDb <- tfModules_withCorr$Target %in% genesInDatabase
# Genes in co-expression modules not available in RcisTargetDatabases:
missingGenes <- sort(unique(tfModules_withCorr[which(!geneInDb),"Target"]))
missingGenes
tfModules_withCorr <- tfModules_withCorr[which(geneInDb),]
    
# Targets with positive correlation
tfModules_Selected <- tfModules_withCorr[which(tfModules_withCorr$corr==1),]

# Add a column with the geneSet name (TF_method)
tfModules_Selected <- cbind(tfModules_Selected, geneSetName=paste(tfModules_Selected$TF, tfModules_Selected$method, sep="_"))
head(tfModules_Selected)

# Split into tfModules (TF-modules, with several methods)
tfModules <- split(tfModules_Selected$Target, tfModules_Selected$geneSetName)

# Keep gene sets with at least 20 genes
tfModules <- tfModules[which(lengths(tfModules)>=20)]

# Add TF to the gene set (used in the following steps, careful if editing)
tfModules <- setNames(lapply(names(tfModules), function(gsn) {
    tf <- strsplit(gsn, "_")[[1]][1]
    unique(c(tf, tfModules[[gsn]]))
    }), names(tfModules))
save(tfModules, file="int/2.1_tfModules_forMotifEnrichmet.RData")
```

```{r}
load("int/2.1_tfModules_forMotifEnrichmet.RData")
tfModulesSummary <- t(sapply(strsplit(names(tfModules), "_"), function(x) x[1:2]))
sort(table(tfModulesSummary[,2]))
```

## Motif enrichment analysis & identifying direct targets
### Motif databases
```{r}
org <- "mm9"
if(org=="hg19")
{
    library(RcisTarget.hg19.motifDatabases.20k)
    
    # Motif rankings (genes x motifs)
    data(hg19_500bpUpstream_motifRanking)
    data(hg19_10kbpAroundTss_motifRanking)
    motifRankings <- list()
    motifRankings[["500bp"]] <- hg19_500bpUpstream_motifRanking
    motifRankings[["10kbp"]] <- hg19_10kbpAroundTss_motifRanking
    
    # Motif annotation (TFs)
    data(hg19_direct_motifAnnotation)
    direct_motifAnnotation <- hg19_direct_motifAnnotation
    data(hg19_inferred_motifAnnotation) # optional
    inferred_motifAnnotation <- hg19_inferred_motifAnnotation
}

if(org=="mm9")
{
    library(RcisTarget.mm9.motifDatabases.20k)
    
    # Motif rankings (genes x motifs)
    data(mm9_500bpUpstream_motifRanking)
    data(mm9_10kbpAroundTss_motifRanking)
    motifRankings <- list()
    motifRankings[["500bp"]] <- mm9_500bpUpstream_motifRanking
    motifRankings[["10kbp"]] <- mm9_10kbpAroundTss_motifRanking
    
    # Motif annotation (TFs)
    data(mm9_direct_motifAnnotation)
    direct_motifAnnotation <- mm9_direct_motifAnnotation
    data(mm9_inferred_motifAnnotation) # optional
    inferred_motifAnnotation <- mm9_inferred_motifAnnotation
}
```

### Run RCisTarget
```{r}
library(RcisTarget)
################################################################
# 1. Calculate motif enrichment for each TF-module

### 1.1 Calculate enrichment
motifs_AUC <- lapply(motifRankings, function(ranking) calcAUC(tfModules, ranking, aucMaxRank=0.01*nrow(ranking@rankings), nCores=4, verbose=FALSE))
save(motifs_AUC, file="int/2.2_motifs_AUC_500bp_10kbp.RData")

### 1.2 Conver to table, filter by NES & add the TFs to which the motif is annotated
# (For each database...)
motifEnrichment <- lapply(motifs_AUC, function(aucOutput)
{
  # Extract the TF of the gene-set name (i.e. MITF_w001):
  tf <- sapply(setNames(strsplit(rownames(aucOutput), "_"), rownames(aucOutput)), function(x) x[[1]])
  
  # Calculate NES and add motif annotation (provide tf in 'highlightTFs'):
  addMotifAnnotation(aucOutput, highlightTFs=tf, nesThreshold=3.0, digits=3,
                  motifAnnot_direct=direct_motifAnnotation,
                  motifAnnot_inferred=inferred_motifAnnotation)
})

# Merge both tables, adding a column that contains the 'motifDb' 
motifEnrichment <- do.call(rbind, lapply(names(motifEnrichment), function(dbName){
  cbind(motifDb=dbName, motifEnrichment[[dbName]])
}))
save(motifEnrichment, file="int/2.3_motifEnrichment.RData")
cat("Number of motifs in the initial enrichment: ", nrow(motifEnrichment))

### 1.3 Keep only the motifs annotated to the initial TF
motifEnrichment_selfMotifs <- motifEnrichment[which(motifEnrichment$TFinDB != ""),, drop=FALSE]
save(motifEnrichment_selfMotifs, file="int/2.4_motifEnrichment_selfMotifs.RData")
cat("Number of motifs annotated to the initial TF: ", nrow(motifEnrichment_selfMotifs))
rm(motifEnrichment)

################################################################
# 2. Prune targets

motifEnrichment_selfMotifs_wGenes <- lapply(names(motifRankings), function(motifDbName){
  addSignificantGenes(resultsTable=motifEnrichment_selfMotifs[motifDb==motifDbName],
                      geneSets=tfModules,
                      rankings=motifRankings[[motifDbName]],
                      maxRank=5000, method="aprox", nCores=4)
  })

library(data.table)
motifEnrichment_selfMotifs_wGenes <- rbindlist(motifEnrichment_selfMotifs_wGenes)
save(motifEnrichment_selfMotifs_wGenes, file="int/2.5_motifEnrichment_selfMotifs_wGenes.RData")

# Save as text:
write.table(motifEnrichment_selfMotifs_wGenes, file="output/Step2_MotifEnrichment.tsv", 
            sep="\t", quote=FALSE, row.names=FALSE)
```

```{r}
load("int/2.5_motifEnrichment_selfMotifs_wGenes.RData")
dim(motifEnrichment_selfMotifs_wGenes)
```

```{r}
motifEnrichment_selfMotifs_wGenes[order(NES,decreasing=TRUE)][1:5,-"enrichedGenes", with=F]
```

### Format regulons
```{r}
library(data.table)
motifEnrichment.asIncidList <- apply(motifEnrichment_selfMotifs_wGenes, 1, function(oneMotifRow) {
  genes <- strsplit(oneMotifRow["enrichedGenes"], ";")[[1]]
  oneMotifRow <- data.frame(rbind(oneMotifRow), stringsAsFactors=FALSE)
  data.frame(oneMotifRow[rep(1, length(genes)),c("NES", "motif", "highlightedTFs", "TFinDB")], genes, stringsAsFactors = FALSE)
})
motifEnrichment.asIncidList <- rbindlist(motifEnrichment.asIncidList)
colnames(motifEnrichment.asIncidList) <- c("NES", "motif", "TF", "annot", "gene")
motifEnrichment.asIncidList <- data.frame(motifEnrichment.asIncidList, stringsAsFactors = FALSE)

# Get targets for each TF, but keep info about best motif/enrichment 
# (directly annotated motifs are considered better)
regulonTargetsInfo <- lapply(split(motifEnrichment.asIncidList, motifEnrichment.asIncidList$TF), function(tfTargets){
  # print(unique(tfTargets$TF))
  tfTable <- as.data.frame(do.call(rbind, lapply(split(tfTargets, tfTargets$gene), function(enrOneGene){
    directAnnot <- "**" %in% enrOneGene$annot
    enrOneGeneByAnnot <- enrOneGene
    if(directAnnot) enrOneGeneByAnnot <- enrOneGeneByAnnot[which(enrOneGene$annot == "**"),]
    bestMotif <- which.max(enrOneGeneByAnnot$NES)
    
    cbind(TF=unique(enrOneGene$TF), gene=unique(enrOneGene$gene), nMotifs=nrow(enrOneGene), 
          bestMotif=as.character(enrOneGeneByAnnot[bestMotif,"motif"]), NES=as.numeric(enrOneGeneByAnnot[bestMotif,"NES"]), 
          directAnnot=directAnnot)
  })), stringsAsFactors=FALSE)
  tfTable[order(tfTable$NES, decreasing = TRUE),]
})
regulonTargetsInfo <- rbindlist(regulonTargetsInfo)
colnames(regulonTargetsInfo) <- c("TF", "gene", "nMotifs", "bestMotif", "NES", "directAnnot")

# Optional: Add Genie3 score
load("int/1.5_GENIE3_linkList.RData")
linkList <- linkList[which(linkList$weight>=0.001),]
rownames(linkList) <- paste(linkList$TF, linkList$Target,sep="__")
regulonTargetsInfo <- cbind(regulonTargetsInfo, Genie3Weight=linkList[paste(regulonTargetsInfo$TF, regulonTargetsInfo$gene,sep="__"),"weight"])

save(regulonTargetsInfo, file="int/2.6_regulonTargetsInfo.RData")
write.table(regulonTargetsInfo, file="output/Step2_regulonTargetsInfo.txt", sep="\t", col.names=TRUE, row.names=FALSE, quote=FALSE)
```

```{r}
regulonTargetsInfo_splitByAnnot <- split(regulonTargetsInfo, regulonTargetsInfo$directAnnot)
regulons <- sapply(split(regulonTargetsInfo_splitByAnnot[["TRUE"]], regulonTargetsInfo_splitByAnnot[["TRUE"]][,"TF"]), function(x) sort(as.character(unlist(x[,"gene"]))))
regulons_extended <- sapply(split(regulonTargetsInfo_splitByAnnot[["FALSE"]],regulonTargetsInfo_splitByAnnot[["FALSE"]][,"TF"]), function(x) unname(x[,"gene"]))
regulons_extended <- sapply(names(regulons_extended), function(tf) sort(unique(c(regulons[[tf]], regulons_extended[[tf]]))))
names(regulons_extended) <- paste(names(regulons_extended), "_extended", sep="")
regulons <- c(regulons, regulons_extended)
save(regulons, file="int/2.6_regulons_asGeneSet.RData")
```

```{r}
load("int/2.6_regulons_asGeneSet.RData")
# Number of regulons and summary of sizes:
length(regulons)
summary(lengths(regulons))
```

### Incidence matrix (i.e. network)
```{r}
incidList <- melt(regulons)
incidMat <- table(incidList[,2], incidList[,1])
save(incidMat, file="int/2.6_regulons_asIncidMat.RData")
dim(incidMat)
```

## Exploring regulons
### How many TFs are self-regulating?
```{r}
table(sapply(names(regulons), function(x) x %in% regulons[[x]]))
```

### Motifs associated to a TF (i.e. MITF regulons):
```{r}
selTF <- "Dlx5"
subsetTable <- motifEnrichment_selfMotifs_wGenes[highlightedTFs %in% selTF][order(NES,decreasing=TRUE)][,-"enrichedGenes", with=F]

subsetTable <- addLogo(subsetTable)
```

```{r}
library(DT)
datatable(subsetTable, escape=FALSE, filter="top", options=list(pageLength=5))
```

### Gene enrichment plots
```{r}
geneSetName <- "Dlx5_top50"
motifDbName <- "10kbp"
selectedMotifs <- subsetTable[geneSet==geneSetName & motifDb==motifDbName, motif]
selectedMotifs <- selectedMotifs[1:3]
```

```{r}
# pdf("int/2.8_RCC_selectedMotifs.pdf")
par(mfrow=c(2,2))
signifGenes_SelectedMotifs <- getSignificantGenes(tfModules[[geneSetName]], 
                                        motifRankings[[motifDbName]],
                                        signifRankingNames=selectedMotifs,
                                        plotCurve=TRUE, maxRank=5000, nCores=4, 
                                        genesFormat="geneList", method="aprox")
```


# Step 3 Analyzing the network activity in each individual cell
## Part 1: Network activity in each cell
### Load regulons & expression matrix
```{r}
sce <- readRDS("data/reads_qc_scran_sc3.rds")
exprMat <- counts(sce)
dim(exprMat)
```

```{r}
load("data/colVars.RData")
cellInfo <- pData(exprMat)[,names(colVars), drop=F]
```

```{r}
cellInfo <- colData(sce)[colnames(exprMat), "sc3_4_clusters", drop = F]
colVars <- list(sc3_4_clusters=setNames(c("forestgreen", "darkorange", "magenta4", "hotpink"), 
                                        c("FAP/MAB", "Schwann cells", "Satellite cells", "Smooth muscle cells")))
save(colVars, file="data/col.Vars.Rdata")
plot.new(); legend(0,1, fill=colVars$sc3_4_clusters, legend=names(colVars$sc3_4_clusters))
```

```{r}
load("int/2.6_regulons_asGeneSet.RData")
regulons <- regulons[order(lengths(regulons), decreasing=TRUE)]
regulons <- regulons[lengths(regulons)>=10]

# Add the TF & rename
regulons <- setNames(lapply(names(regulons), function(tf) sort(unique(c(gsub("_extended", "", tf), regulons[[tf]])))), names(regulons))
names(regulons) <- paste(names(regulons), " (",lengths(regulons), "g)", sep="")
save(regulons, file="int/3.0_regulons_forAUCell.RData")
length(regulons)
```

```{r}
cbind(names(regulons)[1:10])
```

### AUCell

```{r}
library(AUCell)
#library(doMC)
# 1. Create rankings
aucellRankings <- AUCell.buildRankings(exprMat, nCores=4, plotStats=TRUE)
abline(v=aucellRankings@nGenesDetected["1%"], col="skyblue3", lwd=5, lty=3)
save(aucellRankings, file="int/3.1_aucellRankings.RData")

# 2. Calculate AUC
regulonAUC <- AUCell.calcAUC(regulons, aucellRankings, aucMaxRank=aucellRankings@nGenesDetected["1%"], nCores=10)
save(regulonAUC, file="int/3.2_regulonAUC.RData")
```

```{r}
load("int/3.2_regulonAUC.RData")
# Order the modules by similarity, for easier exploration in the upcoming steps & save
variableRegulons <- names(which(apply(getAuc(regulonAUC), 1, sd) > 0))
reguDist <-as.dist(1-cor(t(getAuc(regulonAUC)[variableRegulons,]), method="spear"))
reguClust <- hclust(reguDist, method="ward.D2")
regulonClusters <- setNames(dynamicTreeCut::cutreeDynamic(reguClust, distM=as.matrix(reguDist), verbose = FALSE), reguClust$labels)
regulonOrder <- reguClust$labels[reguClust$order]
regulonOrder <- regulonOrder[order(regulonClusters[regulonOrder], decreasing = TRUE)]
regulonAUC@matrix <- regulonAUC@matrix[regulonOrder,]
save(regulonAUC, file="int/3.2_regulonAUC.RData")

# Export as text (optional):
# write.table(regulonAUC@AUC, file="output/Step3_RegulonActivity_AUC.tsv", sep="\t", quote=FALSE, col.names=NA)
```

#### Overview of cell states according to module activity (tSNE on AUC)

```{r}
library(SCENIC)
# (It is recommended to try different perplexity values)
regulonAUC_subset <- subset(regulonAUC, onlyNonDirectExtended(rownames(regulonAUC)))

# PCA-based t-SNE
set.seed(123)
tsneAUC <- Rtsne::Rtsne(t(getAuc(regulonAUC_subset)), initial_dims=10, perplexity=10)
rownames(tsneAUC$Y) <- colnames(regulonAUC_subset)
colnames(tsneAUC$Y) <- c("tsne1", "tsne2")
save(tsneAUC, file="int/3.3_tsneRegulonAUC_PCA.RData")

# Alternative: Distance-based t-SNE:
corDist <- as.dist(1-cor(getAuc(regulonAUC_subset)))
set.seed(123)
tsneAUC <- Rtsne::Rtsne(corDist, is_distance=TRUE, perplexity=10)
rownames(tsneAUC$Y) <- labels(corDist)
colnames(tsneAUC$Y) <- c("tsne1", "tsne2")
save(tsneAUC, file="int/3.3_tsneRegulonAUC_Dist.RData")
```


```{r}
load("int/3.3_tsneRegulonAUC_PCA.RData")
tSNE <- tsneAUC$Y
par(mfrow=c(1,2))

# Number of genes detected:
nGenesPerCell <- apply(exprMat, 2, function(x) sum(x>0))
colorPal <- grDevices::colorRampPalette(c("darkgreen", "yellow","red"))
cellColorNgenes <- setNames(adjustcolor(colorPal(10), alpha=.8)[as.numeric(cut(nGenesPerCell,breaks=10, right=F,include.lowest=T))], names(nGenesPerCell))

plot(tSNE, col=cellColorNgenes[rownames(tSNE)], pch=16, main="nGenes", sub="t-SNE on regulon activity (AUC)")

# Other known properties:
for(varName in names(colVars))
{
  cellColor <- setNames(colVars[[varName]][cellInfo[,varName]], rownames(cellInfo))
  plot(tSNE, col=cellColor[rownames(tSNE)], pch=16, main=varName, sub="t-SNE on regulon activity (AUC)")
}
```

#### Plot AUC histograms

```{r}
Cairo::CairoPDF("output/Step3_RegulonActivity_AUCtSNE.pdf", width=20, height=5)
par(mfrow=c(1,4))

# tSNE (colored by number of genes detected per cell)
plot(tSNE, col=cellColorNgenes[rownames(tSNE)], pch=16, main="nGenes", sub="t-SNE on regulon activity (AUC)")
plot(tSNE, col=cellColor[rownames(tSNE)], pch=16, main=varName, sub="t-SNE on regulon activity (AUC)")
plot.new(); plot.new()

# Plot module activity, thresholds & assignment:
cells_AUCellThresholds <- plot_aucTsne(tSNE=tSNE, exprMat=exprMat, regulonAUC=regulonAUC, alphaOff=0.1)
dev.off()
save(cells_AUCellThresholds, file="int/3.4_AUCellThresholds.RData")
```

```{r}
load("int/3.4_AUCellThresholds.RData")

# Get cells assigned to each regulon
regulonsCells <- lapply(cells_AUCellThresholds, function(x) x$assignment)

### Save threshold info as text (e.g. to edit/modify...)
trhAssignment <- sapply(cells_AUCellThresholds, function(x) unname(x$aucThr$selected))
commentsThresholds <- sapply(cells_AUCellThresholds, function(x) unname(x$aucThr$comment))

table2edit <- cbind(regulon=names(trhAssignment), 
                    threshold=trhAssignment, 
                    nCellsAssigned=lengths(regulonsCells)[names(trhAssignment)],
                    AUCellComment=commentsThresholds, 
                    nGenes=gsub("[\\(g\\)]", "", regmatches(names(cells_AUCellThresholds), gregexpr("\\(.*?\\)", names(cells_AUCellThresholds)))),
                    clusteringOrder=1:length(trhAssignment), 
                    clusterGroup=regulonClusters[names(trhAssignment)], 
                    onlyNonDirectExtended=(names(trhAssignment) %in% onlyNonDirectExtended(names(trhAssignment))),
                    personalNotes="")
write.table(table2edit, file="int/3.5_1_AUCellThresholds.txt", row.names=F, quote=F, sep="\t")
```


## Part2: Create the binary activity matrix
#### Read the new thresholds
```{r}
manualThresholds <- read.table("int/3.5_2_AUCellThresholds_EDITED.txt", sep="\t", header=T)
newThresholds <- setNames(manualThresholds[,"threshold"], manualThresholds[,"regulon"])
length(newThresholds)
```

```{r}
any(is.na(as.numeric(newThresholds)))
newThresholds  <- setNames(as.numeric(newThresholds), names(newThresholds))
save(newThresholds, file="int/3.5_3_newThresholds.RData")
```

#### Get the cells that have each regulon active with the new thresholds
```{r}
sce <- readRDS("data/reads_qc_scran_sc3.rds")
exprMat <- counts(sce)
load("int/3.5_3_newThresholds.RData")
load("int/3.3_tsneRegulonAUC_PCA.RData"); tSNE <- tsneAUC$Y
load("int/3.2_regulonAUC.RData")

Cairo::CairoPDF("output/Step3_RegulonActivity_AUCtSNE_newThresholds.pdf", width=20, height=5)
par(mfrow=c(1,4))
newAssignment <- plot_aucTsne(exprMat, regulonAUC=regulonAUC, tSNE=tSNE, thresholds=newThresholds)
dev.off()
save(newAssignment, file="int/3.5_4_newAssignment.RData")
```

### Binary regulon activity matrix (Active regulons per cell)
#### Create matrix
```{r}
load("int/3.4_AUCellThresholds.RData")
load("int/3.5_4_newAssignment.RData")
cells_AUCellThresholds <- newAssignment
rm(newAssignment)
```

```{r}
# Get cells assigned to each regulon
regulonsCells <- lapply(cells_AUCellThresholds, function(x) x$assignment)
length(regulonsCells)
```

```{r}
# Conver to matrix (regulons with zero assigned cells are lost)
regulonActivity <- reshape2::melt(regulonsCells)
binaryRegulonActivity <- t(table(regulonActivity[,1], regulonActivity[,2]))
class(binaryRegulonActivity) <- "matrix"
save(binaryRegulonActivity, file="int/3.6_BinaryRegulonActivity.RData")

dim(binaryRegulonActivity)
```

```{r}
binaryRegulonActivity[1:10,1:3]
```

```{r}
binaryRegulonActivity_nonDupl <- binaryRegulonActivity[which(rownames(binaryRegulonActivity) %in% onlyNonDirectExtended(rownames(binaryRegulonActivity))),]
save(binaryRegulonActivity_nonDupl, file="int/3.7_binaryRegulonActivity_nonDupl.RData")
```

#### Matrix overview
```{r}
cbind(nCellsOn=sort(rowSums(binaryRegulonActivity), decreasing=TRUE)[1:15])
```

```{r}
par(mfrow=c(1,2))
boxplot(rowSums(binaryRegulonActivity_nonDupl), main="nCells per regulon", 
        sub='number of cells \nthat have the regulon active',
        col="darkolivegreen1", border="#001100", lwd=2, frame=FALSE)
boxplot(colSums(binaryRegulonActivity_nonDupl), main="nRegulons per Cell", 
        sub='number of regulons \nactive per cell',
        col="darkolivegreen1", border="#001100", lwd=2, frame=FALSE)
```

#### Heatmap
```{r}
load("data/colVars.Rdata")
sce <- readRDS("data/reads_qc_scran_sc3.rds")
exprMat <- counts(sce)
cellInfo <- colData(sce)[,names(colVars), drop=F]
cellInfo$sc3_4_clusters <- revalue(cellInfo$sc3_4_clusters, c("1" = "FAP/MAB", "2" = "Schwann cells", "3" = "Satellite cells", "4" = "Smooth muscle cells"))

minCells <- ncol(sce) * .01
```

```{r}
load("int/3.6_BinaryRegulonActivity.RData")
load("int/3.7_binaryRegulonActivity_nonDupl.RData")

regulonSelection <- list()

# All regulons.
regulonSelection[["All regulons \n (including duplicated regulons)"]] <- rownames(binaryRegulonActivity)

# Active in > 1% cells
regMinCells <- names(which(rowSums(binaryRegulonActivity_nonDupl) > minCells))
regulonSelection[["Regulons active in more than 1% of cells"]] <- regMinCells

# Correlation across regulons (based on binary cell activity)
reguCor <- cor(t(binaryRegulonActivity_nonDupl[regMinCells,]))
diag(reguCor) <- 0

# Regulons that co-ocurr in similar cells. If a regulon is relevant by itself it will not be shown, also check the regulons ignored.
corrRegs <- names(which(rowSums(abs(reguCor) > 0.30) > 0))
regulonSelection[["Regulons with any other regulon correlated\n with abs(cor)>0.30 \n(and active in at least 1% of cells)"]]  <- corrRegs

missingRegs <- rownames(binaryRegulonActivity_nonDupl)[which(!rownames(binaryRegulonActivity_nonDupl) %in% corrRegs)]
regulonSelection[["Regulons no other regulons correlated\n with abs(cor)>0.30 \n or active in fewer than 1% of cells"]]  <- missingRegs

save(regulonSelection,file="int/3.8_regulonSelections.RData")

## Set regulon order (for plotting)
binaryRegulonOrder <- hclust(as.dist(1-reguCor[corrRegs,corrRegs]))
binaryRegulonOrder <- binaryRegulonOrder$labels[binaryRegulonOrder$order]
save(binaryRegulonOrder,file="int/3.9_binaryRegulonOrder.RData")
```

```{r}
for(i in seq_len(length(regulonSelection)))
{
  selRegs <- names(regulonSelection)[i]
  if(length(selRegs)==1)
  {
      binaryMat <- binaryRegulonActivity[regulonSelection[[selRegs]],,drop=FALSE]
      NMF::aheatmap(binaryMat, scale="none", revC=TRUE, main=selRegs,
                    annCol=as.character(cellInfo[colnames(binaryMat),, drop=FALSE]$sc3_4_clusters),
                    annColor=colVars,
                    color = c("white", "black"),
                    filename=paste0("output/Step3.3_binaryRegulonActivity_Heatmap_",i,".pdf"))
   }
}
```

# Step 4: Identifying stable cell states based on the gene-regulatory network activity
## Load cell info and binary activity matrix
```{r}
sce <- readRDS("data/reads_qc_scran_sc3.rds")
exprMat <- counts(sce)
dim(exprMat)
load("data/colVars.Rdata")
cellInfo <- colData(sce)[,names(colVars), drop=F]

#load("int/cellColorNgenes.RData")
nGenesPerCell <- apply(exprMat, 2, function(x) sum(x>0))
colorPal <- grDevices::colorRampPalette(c("darkgreen", "yellow","red"))
cellColorNgenes <- setNames(adjustcolor(colorPal(10), alpha=.8)[as.numeric(cut(nGenesPerCell,breaks=10, right=F,include.lowest=T))], names(nGenesPerCell))

load("int/3.7_binaryRegulonActivity_nonDupl.RData")
tBinaryAct <- t(binaryRegulonActivity_nonDupl)
```

## Calculate t-SNE on the binary regulon activity
```{r}
library(Rtsne)

##################################
# PCA based t-SNE
set.seed(123)
tBinaryAct_jitter <- jitter(tBinaryAct, factor=1)
tsneBinaryActivity_PCA <- Rtsne(tBinaryAct_jitter, initial_dims=5, perplexity=30)
rownames(tsneBinaryActivity_PCA$Y) <- rownames(tBinaryAct_jitter)
colnames(tsneBinaryActivity_PCA$Y) <- c("tsne2", "tsne1")
tsneBinaryActivity_PCA$Y <- tsneBinaryActivity_PCA$Y[,c("tsne1", "tsne2")]
save(tsneBinaryActivity_PCA, file="int/4.1_tsneBinaryActivity_5PC.RData")

##################################
# PCA based t-SNE
set.seed(123)
tBinaryAct_jitter <- jitter(tBinaryAct, factor=1)
tsneBinaryActivity_PCA <- Rtsne(tBinaryAct_jitter, initial_dims=50, perplexity=30)
rownames(tsneBinaryActivity_PCA$Y) <- rownames(tBinaryAct_jitter)
colnames(tsneBinaryActivity_PCA$Y) <- c("tsne2", "tsne1")
tsneBinaryActivity_PCA$Y <- tsneBinaryActivity_PCA$Y[,c("tsne1", "tsne2")]
save(tsneBinaryActivity_PCA, file="int/4.1_tsneBinaryActivity_50PC.RData")

##################################
# Distance-based t-SNE
corDist <- as.dist(1-cor(t(tBinaryAct)))
set.seed(123)
tsneBinaryActivity_Dist <- Rtsne(corDist, is_distance=TRUE, perplexity=30)
rownames(tsneBinaryActivity_Dist$Y) <- labels(corDist)
colnames(tsneBinaryActivity_Dist$Y) <- c("tsne1", "tsne2")
save(tsneBinaryActivity_Dist, file="int/4.1_tsneBinaryActivity_Dist.RData")
```

## Plot t-SNEs
### Coloured by cell properties
```{r}
#library(Cairo)
tSNEs_binary <- list()
load("int/4.1_tsneBinaryActivity_Dist.RData")
tSNEs_binary[["Dist"]] <- tsneBinaryActivity_Dist$Y
load("int/4.1_tsneBinaryActivity_5PC.RData")
tSNEs_binary[["5PC"]] <- tsneBinaryActivity_PCA$Y
load("int/4.1_tsneBinaryActivity_50PC.RData")
tSNEs_binary[["50PC"]] <- tsneBinaryActivity_PCA$Y

for(tsneName in names(tSNEs_binary))
{
    tSNE_binary <- tSNEs_binary[[tsneName]]
    
    # Density
    library(KernSmooth)
    library(RColorBrewer)
    dens2d <- bkde2D(tSNE_binary, 1)$fhat
    
    Cairo::CairoPDF(paste0("output/Step4.1_tsneModuleActivity_",tsneName,".pdf"), width=15, height=5)
    par(mfrow=c(1,3))
    # nGenes
    plot(tSNE_binary, col=cellColorNgenes[rownames(tSNE_binary)], pch=16)
    # density
    image(dens2d, col=brewer.pal(9, "YlOrBr"), axes=FALSE)
    contour(dens2d, add=TRUE, nlevels=5, drawlabels=FALSE)
    
    # Known phenotype:
    if(!is.null(cellInfo))
    {
        nVars <- ncol(cellInfo)
        for(varName in colnames(cellInfo))
        {
            cellColor <- setNames(colVars[[varName]][as.character(cellInfo[,varName])], 
                                  rownames(cellInfo))
            plot(tSNE_binary, col=cellColor[rownames(tSNE_binary)], pch=16, 
                 main=varName, sub="t-SNE on Binary regulon activity",
                 xlab="", ylab="",axes = FALSE)
        }
    }
    # legend(10, 25, names(colVars[[varName]]), fill=colVars[[varName]], cex=.7, bty="n")
    for(i in seq_len(3 - ((nVars+2) %% 3))) # fill remaining slots in page
    {
        plot.new()
    }
    dev.off()
}
```

### Coloured by regulon activity
```{r}
# Choose a t-SNE
load("int/4.1_tsneBinaryActivity_50PC.RData")  
tSNE_binary <- tsneBinaryActivity_PCA$Y
tSNEname <- "tsneBinaryActivity_50PC"

# Load...
load("int/3.6_BinaryRegulonActivity.RData")
load("int/3.9_binaryRegulonOrder.RData")
load("int/3.2_regulonAUC.RData")
load("int/3.4_AUCellThresholds.RData")
```

```{r}
library(SCENIC)
regOrder<- binaryRegulonOrder[which(binaryRegulonOrder %in% rownames(tBinaryAct))]
Cairo::CairoPDF(paste0("output/Step4.2_",tSNEname,"_BinaryRegulons.pdf"), width=20, height=15)
par(mfrow=c(4,6))
cells_trhAssignment <- plot_aucTsne(tSNE=tSNE_binary, exprMat=exprMat,
    regulonAUC=t(tBinaryAct)[binaryRegulonOrder,], cex=1.5, plots="binary", thresholds=0)
dev.off()

regOrder<- binaryRegulonOrder[which(binaryRegulonOrder %in% rownames(regulonAUC))]
Cairo::CairoPDF(paste0("output/Step4.2_",tSNEname,"_AUCRegulons.pdf"), width=20, height=15)
par(mfrow=c(4,6))
cells_trhAssignment <- plot_aucTsne(tSNE=tSNE_binary, exprMat=exprMat, 
    regulonAUC=regulonAUC[regOrder,], cex=1.5, plots="AUC", thresholds=0)
dev.off()

Cairo::CairoPDF(paste0("output/Step4.2_",tSNEname,"_allPlots.pdf"), width=20, height=5)
par(mfrow=c(1,4))
cells_trhAssignment <- plot_aucTsne(tSNE=tSNE_binary, exprMat=exprMat, 
        regulonAUC=regulonAUC[regOrder,],
        alphaOff=0.1, thresholds=cells_AUCellThresholds[regOrder])
dev.off()
```

